Increasing the mortality gap? Social inequality in mortality among elderly and adults in Northern Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}

### 2017-10-24 cape.Rnw
Correcting affiliation
Adding and correcting slides

\author{Sören Edvinsson and Göran Broström \\
Centre for Ageing and Demographic Research \\
Umeå University, Umeå, Sweden}

@@

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1976--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
%  \begin{itemize}
%    \item \emp{Income}
%    \item \emp{Education}
%  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1976 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item \emp{Small} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
\item This suggests an important role of gender and class related expectations on health behaviour
\end{itemize}

@@

\Large

>>>>

\author{Sören Edvinsson and Göran Broström \\
Centre for Demographic and Ageing Research \\
Umeå University, Umeå, Sweden}

@@

\begin{frame}[fragile]{Two time perspectives}

\begin{itemize}
\item Development over \emp{time}.
  \begin{itemize}
\item Explicit or implicit assumption of an \epm{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
\item Antonovsky – divergence followed by convergence
\end{itemize}
\item Development over the \emp/life course}
  \begin{itemize}
\item Convergence, divergence or maintenance.
\item Antonovsky – large differences in adulthood, convergence in old age
\end{itemize}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1801--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1976--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
%  \begin{itemize}
%    \item \emp{Income}
%    \item \emp{Education}
%  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1976 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item \emp{Large} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
\item This suggests an important role of gender and class related expectations on health behaviour
\end{itemize}

@@

\Huge
